---
title: "Permutation Results on CD8+ T Cell CROP-seq Data"
subtitle: "-- In-house scripts"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(grid)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
```

```{r}
source("/project2/xinhe/yifan/Factor_analysis/reports/R/analytics_functions.R")
wkdir <- "/project2/xinhe/yifan/Factor_analysis/Stimulated_T_Cells/"
gsfa_folder <- "gsfa_output_detect_01/all_uncorrected_by_group.use_negctrl/permutation/"
stopifnot(dir.exists(paste0(wkdir, gsfa_folder)))
```

```{r load meta data, fig.width=9}
metadata <- readRDS(paste0(wkdir, "processed_data/metadata.all_T_cells_merged.rds"))
all_KOs.G_mat <- metadata[, 4:24]
KO_names <- colnames(all_KOs.G_mat)
nmarkers <- ncol(all_KOs.G_mat)
```

**We separate the cells into 2 groups:**

Stimulated: 14278 cells;  
Unstimulated stage: 10677 cells.

We modified GSFA so that the associations ($\beta$) between factors and perturbations are estimated for stimulated and unstimulated cells separately.

Here, we summarize the permutation results for GSFA, scMAGeCK, DESeq2, and MAST.

# GSFA Permutation

Within both the stimulated and unstimulated cells, cell labels in the expression data were permuted randomly so that they are no longer correlated with the knock-down conditions. Then GSFA was performed still using all conditions as guides. Factor-guide association as well as the LFSR of each gene were evaluated as usual.

In total, 10 random permutation rounds like this were conducted.

```{r load PM results}
gibbs_list <- list()
lfsr0_list <- list()
lfsr1_list <- list()
for (i in 1:10){
  perm_res <- readRDS(paste0(wkdir, gsfa_folder,
                            "All.gibbs_obj_k20.perm_", i, ".restart.light.rds"))
  gibbs_list[[i]] <- perm_res$posterior_means
  lfsr0_list[[i]] <- perm_res$lfsr0[, -ncol(perm_res$lfsr0)]
  lfsr1_list[[i]] <- perm_res$lfsr1[, -ncol(perm_res$lfsr1)]
}
```

## Factor ~ KO Association

### Posterior Mean of Beta

```{r fig.width=6, fig.height=4}
beta0_pm_list <- list()
for (i in 1:length(gibbs_list)){
  perm_PM <- gibbs_list[[i]]
  beta_pm_matrix <- t(perm_PM$beta0_pm[-(nmarkers + 1), ])
  beta0_pm_list[[i]] <- as.numeric(beta_pm_matrix)
}
p0 <- ggplot(data.frame(beta = unlist(beta0_pm_list)),
             aes(x = beta)) +
  geom_histogram(bins = 30) +
  labs(x = "Estimated association effect size", y = "Count",
       title = "GSFA (Unstimulated)",
       subtitle = "10 Permutations, 21 Targets x 20 Factors") +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 14, hjust = 0.5))

cat(paste("Unstimulated cells, # of |beta|\'s > 0.05:", sum(abs(unlist(beta0_pm_list)) > 0.05)))
cat("\n\n")
```

```{r fig.width=6, fig.height=4}
beta1_pm_list <- list()
for (i in 1:length(gibbs_list)){
  perm_PM <- gibbs_list[[i]]
  beta_pm_matrix <- t(perm_PM$beta1_pm[-(nmarkers + 1), ])
  beta1_pm_list[[i]] <- as.numeric(beta_pm_matrix)
}
p1 <- ggplot(data.frame(beta = unlist(beta1_pm_list)),
             aes(x = beta)) +
  geom_histogram(bins = 30) +
  labs(x = "Estimated association effect size", y = "Count",
       title = "GSFA (Stimulated)",
       subtitle = "10 Permutations, 21 Targets x 20 Factors") +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 14, hjust = 0.5))
cat(paste("Stimulated cells, # of |beta|\'s > 0.05:", sum(abs(unlist(beta1_pm_list)) > 0.05)))
cat("\n\n")
```

```{r Fig_S4A, fig.width=9, fig.height=4}
grid.arrange(p1, p0, nrow = 1)
```

### Beta PIP
```{r fig.width=6, fig.height=4}
beta0_pip_list <- list()
for (i in 1:length(gibbs_list)){
  perm_PM <- gibbs_list[[i]]
  beta_pip_matrix <- t(perm_PM$Gamma0_pm[-(nmarkers + 1), ])
  beta0_pip_list[[i]] <- as.numeric(beta_pip_matrix)
}
p0 <- ggplot(data.frame(pip = unlist(beta0_pip_list)),
             aes(x = pip)) +
  geom_histogram(bins = 30) +
  labs(x = "Estimated PIP of association", y = "Count",
       title = "GSFA (Unstimulated)",
       subtitle = "10 Permutations, 21 Targets x 20 Factors") +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 14, hjust = 0.5))

cat(paste("Unstimulated cells, # of PIPs > 0.8:", sum(unlist(beta0_pip_list) > 0.8)))
cat("\n\n")
```

```{r fig.width=6, fig.height=4}
beta1_pip_list <- list()
for (i in 1:length(gibbs_list)){
  perm_PM <- gibbs_list[[i]]
  beta_pip_matrix <- t(perm_PM$Gamma1_pm[-(nmarkers + 1), ])
  beta1_pip_list[[i]] <- as.numeric(beta_pip_matrix)
}
p1 <- ggplot(data.frame(pip = unlist(beta1_pip_list)),
             aes(x = pip)) +
  geom_histogram(bins = 30) +
  labs(x = "Estimated PIP of association", y = "Count",
       title = "GSFA (Stimulated)",
       subtitle = "10 Permutations, 21 Targets x 20 Factors") +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 14, hjust = 0.5))

cat(paste("Stimulated cells, # of PIPs > 0.8:", sum(unlist(beta1_pip_list) > 0.8)))
cat("\n\n")
```

```{r Fig_S4B, fig.width=9, fig.height=4}
grid.arrange(p1, p0, nrow = 1)
```

## LFSR

```{r plot LFSR group0 results}
lfsr0_df <- data.frame(lfsr = unlist(lfsr0_list),
                       type = "Unstimulated")
percent_str0 <- scales::percent(sum(lfsr0_df$lfsr < 0.05) /
                                  nrow(lfsr0_df), accuracy = 0.1)
grob0 <- grobTree(textGrob(paste(percent_str0, "data < 0.05"),
                                 x = 0.1,  y = 0.1, hjust = 0,
                           gp = gpar(col = "red", fontsize = 12)))
p0 <- ggplot(lfsr0_df, aes(x = lfsr)) +
  geom_histogram(bins = 30) +
  labs(x = "LFSR", y = "Count",
       title = "GSFA (Unstimulated)",
       subtitle = "10 Permutations, 21 Targets x 6000 Genes") +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5)) +
  annotation_custom(grob0)
```

```{r plot LFSR group1 results}
lfsr1_df <- data.frame(lfsr = unlist(lfsr1_list),
                       type = "Stimulated")
percent_str1 <- scales::percent(sum(lfsr1_df$lfsr < 0.05) /
                                  nrow(lfsr1_df), accuracy = 0.1)
grob1 <- grobTree(textGrob(paste(percent_str1, "data < 0.05"),
                           x = 0.1,  y = 0.1, hjust = 0,
                           gp = gpar(col = "red", fontsize = 12)))
p1 <- ggplot(lfsr1_df, aes(x = lfsr)) +
  geom_histogram(bins = 30) +
  labs(x = "LFSR", y = "Count",
       title = "GSFA (Stimulated)",
       subtitle = "10 Permutations, 21 Targets x 6000 Genes") +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5)) +
  annotation_custom(grob1)
```


```{r Fig_S4C, fig.width=9, fig.height=4}
grid.arrange(p1, p0, nrow = 1)
```

# scMAGeCK Permutation (Stimulated)

Within stimulated cells, cell labels in the expression data were permuted randomly so that they are no longer correlated with the knock-down conditions. Then scMAGeCK-LR test was performed for all guides at once.

In total, 10 random permutation rounds like this were conducted.

**The outputs are empirical p values**, some of them equal to 0 exactly, and had to be replaced with 0.0005 for QQ-plot.

```{r load scmageck results}
scmageck_list <- list()
for (i in 1:10){
  res <- readRDS(paste0(wkdir, "scmageck/scmageck_lr.TCells_stim.perm_", i, ".rds"))
  res <- reshape2::melt(res$pval, variable.name = "target", value.name = "pvalue")
  res$pvalue[res$pvalue==0] <- 0.0005 # minimum non-zero value is 0.001
  scmageck_list[[i]] <- res
}
```

```{r fig.width=10, fig.height=5, eval=F, include=F}
for (i in 1:10){
  summ_pvalues(scmageck_list[[i]]$pvalue,
               title_text = paste0("Stimulated Cells\n",
                                   "Permutation ", i, ", 6000 genes x 21 Targets"))
}
```

## Combined from 10 permutations

```{r Fig_S4D, fig.width=5, fig.height=5}
qqunif.plot(do.call(rbind, scmageck_list)$pvalue,
            main = paste0("scMAGeCK Permutations (Stimulated)"))
```

## Original scMAGeCK result

```{r}
scmageck_unpermuted <- readRDS(paste0(wkdir, "scmageck/scmageck_lr.TCells_stim.dev_res_top_6k.rds"))
scmageck_res <- reshape2::melt(scmageck_unpermuted$pval, variable.name = "target", value.name = "pvalue")
scmageck_res$pvalue[scmageck_res$pvalue==0] <- 0.0005
```

```{r fig.width=10, fig.height=5}
summ_pvalues(scmageck_res$pvalue,
             title_text = paste0("scMAGeCK (Stimulated)\n",
                                 "Original, 6000 genes x 21 Targets"))
```

# DESeq2 Permutation (Stimulated)

Within stimulated cells, cell labels in the expression data were permuted randomly so that they are no longer correlated with the knock-down conditions. Then DESeq2 test was performed under each guide.

In total, 10 random permutation rounds like this were conducted.

```{r load deseq2 results}
deseq_list <- list()
for (i in 1:10){
  perm_lst <- list()
  for (m in KO_names){
    res <- readRDS(paste0(wkdir, "processed_data/DESeq2/all_by_stim_dev_6k/perm_",
                            i, "/gRNA_", m, ".dev_res_top6k.rds"))
    res <- as.data.frame(res@listData, row.names = res@rownames)
    res$target <- m
    perm_lst[[m]] <- res
  }
  deseq_list[[i]] <- do.call(rbind, perm_lst)
}
```

```{r fig.width=10, fig.height=5, eval=F, include=F}
for (i in 1:10){
  summ_pvalues(deseq_list[[i]]$pvalue,
             title_text = paste0("Stimulated Cells\n",
                                 "Permutation ", i, ", 6000 genes x 21 Targets"))
}
```

## Combined from 10 permutations

```{r Fig_S4E, fig.width=5, fig.height=5}
qqunif.plot(do.call(rbind, deseq_list)$pvalue,
            main = paste0("DESeq2 Permutations (Stimulated)"))
```

## Original DESeq2 DGE result

```{r}
deseq_unpermuted <- list()
for (m in KO_names){
  fname <- paste0(wkdir, "processed_data/DESeq2/all_by_stim_dev_6k/gRNA_", m, ".dev_res_top6k.rds")
  res <- readRDS(fname)
  res <- as.data.frame(res@listData, row.names = res@rownames)
  res$target <- m
  deseq_unpermuted[[m]] <- res
}
deseq_unpermuted <- do.call(rbind, deseq_unpermuted)
```

```{r fig.width=10, fig.height=5}
summ_pvalues(deseq_unpermuted$pvalue,
             title_text = paste0("DESeq2 (Stimulated)\n",
                                 "Original, 6000 genes x 21 Targets"))
```

# MAST Permutation (Stimulated)

Within stimulated cells, cell labels in the expression data were permuted randomly so that they are no longer correlated with the knock-down conditions. Then MAST likelihood ratio test was performed under each guide.

In total, 10 random permutation rounds like this were conducted.

```{r load MAST results}
mast_list <- list()
for (i in 1:10){
  perm_lst <- list()
  for (m in KO_names){
    res <- readRDS(paste0(wkdir, "processed_data/MAST/all_by_stim_dev_6k/perm_",
                            i, "/gRNA_", m, ".dev_res_top6k.rds"))
    res$target <- m
    perm_lst[[m]] <- res
  }
  mast_list[[i]] <- do.call(rbind, perm_lst)
}
```

```{r fig.width=10, fig.height=5, eval=F, include=F}
for (i in 1:10){
  summ_pvalues(mast_list[[i]]$pval,
             title_text = paste0("Stimulated Cells\n",
                                 "Permutation ", i, ", 6000 genes x 21 Targets"))
}
```

## Combined from 10 permutations

```{r Fig_S4F, fig.width=5, fig.height=5}
qqunif.plot(do.call(rbind, mast_list)$pval,
            main = paste0("MAST Permutations (Stimulated)"))
```

## Original MAST DGE result

```{r}
mast_unpermuted <- list()
for (m in KO_names){
  fname <- paste0(wkdir, "processed_data/MAST/all_by_stim_dev_6k/gRNA_", m, ".dev_res_top6k.rds")
  res <- readRDS(fname)
  res$target <- m
  mast_unpermuted[[m]] <- res
}
mast_unpermuted <- do.call(rbind, mast_unpermuted)
```

```{r fig.width=10, fig.height=5}
summ_pvalues(mast_unpermuted$pval,
             title_text = paste0("MAST (Stimulated)\n",
                                 "Original, 6000 genes x 21 Targets"))
```

# edgeR Permutation (Stimulated)

Within stimulated cells, cell labels in the expression data were permuted randomly so that they are no longer correlated with the knock-down conditions. Then edgeR QLF test was performed under each guide.

In total, 10 random permutation rounds like this were conducted.

```{r load edgeR results}
edger_list <- list()
for (i in 1:10){
  perm_lst <- list()
  for (m in KO_names){
    res <- readRDS(paste0(wkdir, "processed_data/edgeR_qlf/all_by_stim_dev_6k/perm_",
                            i, "/gRNA_", m, ".dev_res_top6k.rds"))
    res$target <- m
    perm_lst[[m]] <- res
  }
  edger_list[[i]] <- do.call(rbind, perm_lst)
}
```

```{r fig.width=10, fig.height=5, eval=F, include=F}
for (i in 1:10){
  summ_pvalues(edger_list[[i]]$PValue,
             title_text = paste0("Stimulated Cells\n",
                                 "Permutation ", i, ", 6000 genes x 21 Targets"))
}
```

## Combined from 10 permutations

```{r Fig_S4H, fig.width=5, fig.height=5}
qqunif.plot(do.call(rbind, edger_list)$PValue,
            main = paste0("edgeR-QLF Permutations (Stimulated)"))
```

## Original edgeR DGE result

```{r}
edger_unpermuted <- list()
for (m in KO_names){
  fname <- paste0(wkdir, "processed_data/edgeR_qlf/all_by_stim_dev_6k/gRNA_", m, ".dev_res_top6k.rds")
  res <- readRDS(fname)
  res$target <- m
  edger_unpermuted[[m]] <- res
}
edger_unpermuted <- do.call(rbind, edger_unpermuted)
```

```{r fig.width=10, fig.height=5}
summ_pvalues(edger_unpermuted$PValue,
             title_text = paste0("edgeR-QLF (Stimulated)\n",
                                 "Original, 6000 genes x 21 Targets"))
```

